---
title: "PhillyStat360 - OVS Exploratory Analysis & Parcel Drill-Down"
author: "MUSA Practicum Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Setup

```{r}
library(tidyverse)
library(janitor)
library(knitr)
library(scales)
library(patchwork)
library(sf)
library(tigris)

data_path <- "../rawdata/"
```


## 1. Load Data

```{r read-data, cache=TRUE}
# Primary table: all Philadelphia parcels
properties <- read_csv(paste0(data_path, "opa_properties_public.csv"),
                       show_col_types = FALSE) %>% clean_names()

# Three core OVS sources
clean_seal <- read_csv(paste0(data_path, "clean_seal.csv"),
                       show_col_types = FALSE) %>% clean_names()

violations <- read_csv(paste0(data_path, "violations.csv"),
                       show_col_types = FALSE) %>% clean_names()

licenses <- read_csv(paste0(data_path, "business_licenses.csv"),
                     show_col_types = FALSE) %>% clean_names()

permits <- read_csv(paste0(data_path, "permits.csv"),
                   show_col_types = FALSE) %>% clean_names()

# Supplementary tables
unsafe <- read_csv(paste0(data_path, "unsafe.csv"),
                   show_col_types = FALSE) %>% clean_names()

imm_dang <- read_csv(paste0(data_path, "imm_dang.csv"),
                     show_col_types = FALSE) %>% clean_names()

cat("properties:", comma(nrow(properties)), "rows\n")
cat("clean_seal:", comma(nrow(clean_seal)), "rows\n")
cat("violations:", comma(nrow(violations)), "rows\n")
cat("licenses:", comma(nrow(licenses)), "rows\n")
cat("unsafe:", comma(nrow(unsafe)), "rows\n")
cat("imm_dang:", comma(nrow(imm_dang)), "rows\n")
```

```{r pwd parcels}
#adding in pwd parcels instead (seen as the gold standard for parcels and also has geog)
parcel_sf <- st_read(
  paste0(data_path,"PWD_PARCELS.geojson"),quiet = TRUE) %>% clean_names()
cat("parcel_sf:", comma(nrow(parcel_sf)), "rows\n")

```


## 2. Constructing the Observed Vacancy Status (OVS)

Based on the VacancyGuide2026 methodology and client kickoff meeting, a parcel is classified as OVS=1 (observed vacant) if it meets **any** of the following criteria:

- **Source 1: Clean & Seal** — Property appears in the L+I Clean & Seal records (government-sealed dangerous building)
- **Source 2: Vacant Violations** — Property has specific vacancy-related violation codes with status = 'OPEN'
- **Source 3: Vacant License** — Property holds an active Vacant Property license

#violations
violationcode IN ('PM15-304.10', 'PM15-304.11', 'PM15-304.19', 'PM15-304.19O', 'PM15-304.2', 'PM15-304.8', 'PM15-308.1', 'PM15-309.1', 'PM15-309.3', 'PM15-309.4', 'PM-302.2/1', 'PM-302.2/2', 'PM-302.3/1', 'PM-302.4/6', 'PM-302.4/7', 'PM-304.1/1', 'PM-304.6/1', 'PM-304.8/1', 'PM-304.8/15', 'PM-304.8/16', 'PM-304.8/17', 'PM-304.8/18', 'PM-304.8/19', 'PM-304.8/2', 'PM-304.8/20', 'PM-304.8/21')
```{r}
#target violation codes
target_codes <- c(
  'PM15-304.10', 'PM15-304.11', 'PM15-304.19', 'PM15-304.19O', 'PM15-304.2', 'PM15-304.8', 'PM15-308.1', 'PM15-309.1', 'PM15-309.3', 'PM15-309.4', 'PM-302.2/1', 'PM-302.2/2', 'PM-302.3/1', 'PM-302.4/6', 'PM-302.4/7', 'PM-304.1/1', 'PM-304.6/1', 'PM-304.8/1', 'PM-304.8/15', 'PM-304.8/16', 'PM-304.8/17', 'PM-304.8/18', 'PM-304.8/19', 'PM-304.8/2', 'PM-304.8/20', 'PM-304.8/21'
)

#Check for open violations from the list
violation_scores <- violations %>%
  filter(
    violationcode %in% target_codes,
    tolower(violationstatus) == 'open'
  ) %>%
  group_by(opa_account_num) %>%
  summarise(violation_points = 1) %>%
  ungroup()

```

('9-3905', 'CP-103', 'PM-102.4/2', 'PM-102.4/4', 'PM15-108.1', 'PM15-108.2', 'PM15-109.1', 'PM15-110.1', 'PM15-301', 'PM15-304.1', 'PM15-304.19V', 'PM15-304.1A', 'PM15-304.1B', 'PM15-304.1C', 'PM15-304.1D', 'PM15-304.1E', 'PM15-304.1F', 'PM15-304.1G', 'PM15-304.1H', 'PM15-304.1I', 'PM15-304.1J', 'PM15-901.1', 'PM15-901.2', 'PM15-902.1', 'PM-304.4/1', 'PM-306.0/1', 'PM-306.0/90', 'PM-306.2/1', 'PM-306.2/4', 'PM-306.2/90', 'PM-306.2/91', 'PM-306.2/92', 'PM-307.1/1', 'PM-307.1/12', 'PM-307.1/13', 'PM-307.1/19', 'PM-307.1/3', 'PM-307.1/4', 'PM-307.1/5', 'PM-307.1/6', 'PM-307.1/7', 'PM-307.1/8', 'PM-307.1/9', 'PM-307.1/90', 'PM-307.2/1', 'PM-307.3/1', 'PM-307.3/2', 'PM-307.3/90', 'PM-307.3/91', 'PM-308.1/1', 'PM-308.1/10', 'PM-308.1/11', 'PM-308.1/12', 'PM-308.1/2', 'PM-308.1/3', 'PM-308.1/4', 'PM-308.1/5', 'PM-308.1/6', 'PM-308.1/7', 'PM-308.1/8', 'PM-308.1/9', ‘PM-307.1/2’, ‘PM-307.1/11')

```{r}
target_codes <- c(
'9-3905', 'CP-103', 'PM-102.4/2', 'PM-102.4/4', 'PM15-108.1', 'PM15-108.2', 'PM15-109.1', 'PM15-110.1', 'PM15-301', 'PM15-304.1', 'PM15-304.19V', 'PM15-304.1A', 'PM15-304.1B', 'PM15-304.1C', 'PM15-304.1D', 'PM15-304.1E', 'PM15-304.1F', 'PM15-304.1G', 'PM15-304.1H', 'PM15-304.1I', 'PM15-304.1J', 'PM15-901.1', 'PM15-901.2', 'PM15-902.1', 'PM-304.4/1', 'PM-306.0/1', 'PM-306.0/90', 'PM-306.2/1', 'PM-306.2/4', 'PM-306.2/90', 'PM-306.2/91', 'PM-306.2/92', 'PM-307.1/1', 'PM-307.1/12', 'PM-307.1/13', 'PM-307.1/19', 'PM-307.1/3', 'PM-307.1/4', 'PM-307.1/5', 'PM-307.1/6', 'PM-307.1/7', 'PM-307.1/8', 'PM-307.1/9', 'PM-307.1/90', 'PM-307.2/1', 'PM-307.3/1', 'PM-307.3/2', 'PM-307.3/90', 'PM-307.3/91', 'PM-308.1/1', 'PM-308.1/10', 'PM-308.1/11', 'PM-308.1/12', 'PM-308.1/2', 'PM-308.1/3', 'PM-308.1/4', 'PM-308.1/5', 'PM-308.1/6', 'PM-308.1/7', 'PM-308.1/8', 'PM-308.1/9', 'PM-307.1/2', 'PM-307.1/11'
)
```



#Permits
```{r}
permit_summary <- permits %>%
  filter(!is.na(permitcompleteddate)) %>%
  group_by(property_id) %>%
  summarise(
    last_demo_date = max(permitcompleteddate[str_detect(tolower(permittype), "demolition")], na.rm = TRUE),
    last_new_con_date = max(permitcompleteddate[str_detect(tolower(permittype), "new construction")], na.rm = TRUE)
  ) %>%
  mutate(
    last_demo_date = if_else(is.infinite(last_demo_date), as.Date(NA), last_demo_date),
    last_new_con_date = if_else(is.infinite(last_new_con_date), as.Date(NA), last_new_con_date)
  )
```












______

### 2a. Source 1: Clean & Seal Parcels

```{r ovs-cleanseal}
cs_parcels <- clean_seal %>%
  filter(!is.na(opa_account_num)) %>%
  distinct(opa_account_num) %>%
  mutate(ovs_clean_seal = 1)

cat("Distinct parcels in Clean & Seal:", comma(nrow(cs_parcels)), "\n")
```

### 2b. Source 2: Vacant Violations (Currently OPEN)

Per VacancyGuide, only violations with `violationstatus = 'OPEN'` are considered.

```{r ovs-violations}
# Vacancy-related violation codes per VacancyGuide2026
vacant_viol_codes <- c(
  "9-3904",     # Vacant Lot License
  "9-3905",     # Vacant Structure License
  "PM15-901.1", # Vacant & Open Properties
  "PM15-901.2", # Vacant Properties - Blighted Condition
  "PM15-301",   # Vacant Structure & Land
  "CP-102",     # License - Vacant Lot
  "CP-103",     # License - Vacant Building
  "PM-102.4/1", # License - Vac Lot
  "PM-102.4/2", # License - Vac Res Bldg
  "PM-102.4/3", # License - Vac Wharf/Pier
  "PM-102.4/4"  # License - Vac Comm Bldg
)

viol_parcels <- violations %>%
  filter(violationcode %in% vacant_viol_codes,
         tolower(violationstatus) == "open") %>%
  distinct(opa_account_num) %>%
  filter(!is.na(opa_account_num)) %>%
  mutate(ovs_violation = 1)

cat("Distinct parcels with OPEN vacant violations:", comma(nrow(viol_parcels)), "\n")

# Distribution of OPEN vacant violation codes
violations %>%
  filter(violationcode %in% vacant_viol_codes,
         tolower(violationstatus) == "open") %>%
  count(violationcode, sort = TRUE) %>%
  kable(caption = "Distribution of Currently OPEN Vacant Violation Codes")
```

### 2c. Source 3: Active Vacant License

```{r ovs-license}
# Per VacancyGuide: licensestatus = 'Active' AND revenuecode IN ('3219', '3634')
# 3219 = Residential Vacant License, 3634 = Vacant Commercial Property
vacant_license_parcels <- licenses %>%
  filter(tolower(licensestatus) == "active",
         revenuecode %in% c("3219")) %>%
  distinct(opa_account_num) %>%
  filter(!is.na(opa_account_num)) %>%
  mutate(ovs_license = 1)

cat("Distinct parcels with Active Vacant License:", comma(nrow(vacant_license_parcels)), "\n")
```

### 2d. Merging OVS onto Properties

```{r ovs-merge}
ovs_df <- properties %>%
  select(parcel_number, location, owner_1, category_code, category_code_description,
         year_built, total_livable_area, market_value, sale_date, sale_price,
         exterior_condition, interior_condition, zoning, zip_code, census_tract) %>%
  left_join(cs_parcels, by = c("parcel_number" = "opa_account_num")) %>%
  left_join(viol_parcels, by = c("parcel_number" = "opa_account_num")) %>%
  left_join(vacant_license_parcels, by = c("parcel_number" = "opa_account_num")) %>%
  replace_na(list(ovs_clean_seal = 0, ovs_violation = 0, ovs_license = 0)) %>%
  mutate(
    ovs = if_else(ovs_clean_seal == 1 | ovs_violation == 1 | ovs_license == 1, 1, 0),
    ovs_sources = ovs_clean_seal + ovs_violation + ovs_license
  )

cat("Total parcels:", comma(nrow(ovs_df)), "\n")
cat("OVS=1 (Observed Vacant):", comma(sum(ovs_df$ovs == 1)), "\n")
cat("OVS=0:", comma(sum(ovs_df$ovs == 0)), "\n")
cat("Vacancy rate:", percent(mean(ovs_df$ovs), accuracy = 0.1), "\n")
```

## 3. Exploratory Analysis Around OVS

#cleaning data

only want single family(1), apartments>4units(14), multi-family (2), mixed use(3). the rest of them are irrelevant to our analysis on properties. 

```{r}
ovs_df <- ovs_df %>% 
  filter(category_code %in% c("1","2","3","14"))
```

inner join
```{r}
ovs_df <- ovs_df %>% 
  left_join(
  parcel_sf %>% select(brt_id, bldg_desc, shape_area, shape_length, geometry), by = c("parcel_number" = "brt_id")
) %>% 
  st_as_sf()
  
#glimpse(ovs_spatial)
```


```{r}
cat("Total parcels:", comma(nrow(ovs_df)), "\n")
cat("OVS=1 (Observed Vacant):", comma(sum(ovs_df$ovs == 1)), "\n")
cat("OVS=0:", comma(sum(ovs_df$ovs == 0)), "\n")
cat("Vacancy rate:", percent(mean(ovs_df$ovs), accuracy = 0.1), "\n")
```


### 3a. OVS Source Overlap

How much do the three sources agree? Are parcels typically flagged by one source or multiple?

```{r ovs-overlap}
ovs_positive <- ovs_df %>% filter(ovs == 1)

cat("=== OVS=1 Source Breakdown ===\n")
cat("Clean & Seal only:", comma(sum(ovs_positive$ovs_clean_seal == 1 & ovs_positive$ovs_violation == 0 & ovs_positive$ovs_license == 0)), "\n")
cat("Violation only:", comma(sum(ovs_positive$ovs_clean_seal == 0 & ovs_positive$ovs_violation == 1 & ovs_positive$ovs_license == 0)), "\n")
cat("License only:", comma(sum(ovs_positive$ovs_clean_seal == 0 & ovs_positive$ovs_violation == 0 & ovs_positive$ovs_license == 1)), "\n")
cat("Two sources overlap:", comma(sum(ovs_positive$ovs_sources == 2)), "\n")
cat("All three sources:", comma(sum(ovs_positive$ovs_sources == 3)), "\n")
```

```{r ovs-overlap-plot, fig.width=8, fig.height=5}
ovs_combo <- ovs_positive %>%
  mutate(source_combo = case_when(
    ovs_clean_seal == 1 & ovs_violation == 1 & ovs_license == 1 ~ "All Three",
    ovs_clean_seal == 1 & ovs_violation == 1 ~ "Clean&Seal + Violation",
    ovs_clean_seal == 1 & ovs_license == 1 ~ "Clean&Seal + License",
    ovs_violation == 1 & ovs_license == 1 ~ "Violation + License",
    ovs_clean_seal == 1 ~ "Clean & Seal Only",
    ovs_violation == 1 ~ "Violation Only",
    ovs_license == 1 ~ "License Only"
  )) %>%
  count(source_combo, sort = TRUE) %>%
  ggplot(aes(x = reorder(source_combo, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "OVS=1 Parcels by Source Combination",
       x = NULL, y = "Number of Parcels") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))

ovs_combo

ggsave("ovs_combination_plot.png", plot = ovs_combo, width = 10, height = 6, dpi = 300)
```

### 3b. OVS vs. Property Characteristics

```{r ovs-features, fig.width=10, fig.height=8}
# Market Value comparison (log scale due to extreme skew)
#p1 <- ovs_df %>%
#  filter(market_value > 0) %>%
#  ggplot(aes(x = factor(ovs, labels = c("Not Vacant", "Vacant")),
#             y = market_value)) +
#  geom_boxplot(fill = c("gray80", "tomato"), outlier.alpha = 0.05) +
#  scale_y_log10(labels = dollar) +
#  labs(title = "Market Value by OVS", x = NULL, y = "Market Value (log scale)") +
#  theme_minimal()#

# Year Built comparison
#p2 <- ovs_df %>%
#  filter(!is.na(year_built), year_built > 1700) %>%
#  ggplot(aes(x = year_built, fill = factor(ovs, labels = c("Not Vacant", "Vacant")))) +
#  geom_density(alpha = 0.5) +
#  labs(title = "Year Built Distribution by OVS",
#       x = "Year Built", y = "Density", fill = NULL) +
#  theme_minimal()

# Exterior Condition comparison
#p3 <- ovs_df %>%
#  filter(!is.na(exterior_condition)) %>%
#  count(ovs, exterior_condition) %>%
#  group_by(ovs) %>%
#  mutate(pct = n / sum(n)) %>%
#  ggplot(aes(x = factor(exterior_condition), y = pct,
#             fill = factor(ovs, labels = c("Not Vacant", "Vacant")))) +
#  geom_col(position = "dodge") +
#  scale_y_continuous(labels = percent) +
#  labs(title = "Exterior Condition Distribution by OVS",
#       subtitle = "1 = Best, 7 = Worst",
#       x = "Exterior Condition", y = "Proportion", fill = NULL) +
#  theme_minimal()#

# OPA Category Code = 6 (Vacant Land) rate
#p4 <- ovs_df %>%
#  mutate(is_vacant_land = category_code == "6") %>%
#  count(ovs, is_vacant_land) %>%
#  group_by(ovs) %>%
# mutate(pct = n / sum(n)) %>%
# filter(is_vacant_land == TRUE) %>%
#  ggplot(aes(x = factor(ovs, labels = c("Not Vacant", "Vacant")), y = pct)) +
#  geom_col(fill = c("gray60", "tomato"), width = 0.5) +
#  scale_y_continuous(labels = percent) +
#  labs(title = "OPA Category Code = 6 (Vacant Land) Rate",
#       x = NULL, y = "% with Vacant Land Code") +
#  theme_minimal()

#(p1 | p2) / (p3 | p4)

#ggsave("market_ovs.png", plot = p1, width = 10, height = 6, dpi = 300)
#ggsave("year_built.png", plot = p2, width = 10, height = 6, dpi = 300)
##ggsave("exteriorcondition.png", plot = p3, width = 10, height = 6, dpi = 300)
#ggsave("opa_cat.png", plot = p4, width = 10, height = 6, dpi = 300)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(sf) # Ensure sf is loaded to use st_drop_geometry

condition_dist_plot <- ovs_df %>%
  # 1. STRIP GEOMETRY: This is the most likely reason your R is freezing. 
  # Bar charts don't need coordinates.
  st_drop_geometry() %>% 
  
  # 2. Select and ensure Numeric
  select(ovs, exterior_condition, interior_condition) %>%
  mutate(across(c(exterior_condition, interior_condition), as.numeric)) %>%
  
  # 3. Pivot the data
  pivot_longer(
    cols = c(exterior_condition, interior_condition),
    names_to = "condition_type",
    values_to = "score"
  ) %>%
  
  # 4. Filter and Count (Aggregating here makes the plot render instantly)
  filter(!is.na(score)) %>%
  count(ovs, condition_type, score) %>%
  group_by(ovs, condition_type) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  
  # 5. Clean labels
  mutate(condition_type = if_else(
    condition_type == "exterior_condition", 
    "Exterior Condition", 
    "Interior Condition"
  )) %>%
  
  # 6. Plot
  ggplot(aes(x = factor(score), y = pct,
             fill = factor(ovs, labels = c("Not Vacant", "Vacant")))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent) +
  facet_wrap(~ condition_type, ncol = 1) + 
  labs(title = "Condition Distribution by OVS",
       subtitle = "1 = Best, 7 = Worst",
       x = "Condition Score", y = "Proportion", fill = NULL) +
  theme_minimal()

ggsave("condition_distribution.png", 
       plot = condition_dist_plot, 
       width = 10, 
       height = 8, 
       dpi = 300)
```


#spatial

```{r}
vacant_parcels_map <- ovs_df %>%
  filter(!st_is_empty(geometry)) %>% 
  st_centroid() %>% 
  ggplot() +
  geom_sf(aes(
    color = factor(ovs), 
    size = factor(ovs), 
    alpha = factor(ovs)
  )) +
  scale_color_manual(
    values = c("0" = "gray80", "1" = "red"),
    labels = c("Occupied / Unknown", "Vacant")
  ) +
  scale_size_manual(
    values = c("0" = 0.1, "1" = 0.8), 
    guide = "none" # Hide from legend to keep it clean
  ) +
  # 4. Make Vacant dots fully opaque and Occupied dots very faint
  scale_alpha_manual(
    values = c("0" = 0.2, "1" = 1.0), 
    guide = "none"
  ) +
  theme_void() +
  labs(title = "All Parcels: Vacant vs Occupied", color = "Status") +
  guides(color = guide_legend(override.aes = list(size = 3)))

#ggsave("vacant_parcels_map.png", plot = vacant_parcels_map, width = 10, height = 10, dpi = 300)
```


```{r}
vacant_parcels_map
```

```{r}
vacant_only_map <- ovs_df %>%
  filter(!st_is_empty(geometry), ovs == 1) %>% 
  st_centroid() %>% 
  ggplot() +
  geom_sf(color = "red", size = 0.05, alpha = 0.4) +
  theme_void() +
  labs(title = "Vacant Parcels Only")
vacant_only_map
ggsave("vacant_only.png", plot = vacant_only_map, width = 10, height = 10)
```

```{r}
occupied_only_map <- ovs_df %>%
  filter(!st_is_empty(geometry), ovs == 0) %>% 
  st_centroid() %>% 
  ggplot() +
  geom_sf(color = "gray80", size = 0.1, alpha = 0.3) +
  theme_void() +
  labs(title = "Occupied / Unknown Parcels Only")
occupied_only_map
ggsave("occupied_only.png", plot = occupied_only_map, width = 10, height = 10)
```

### 3c. Spatial Distribution of Vacancy (by Zip Code)

#```{r ovs-spatial, fig.width=10, fig.height=6}
zip_summary <- ovs_df %>%
  filter(!is.na(zip_code)) %>%
  group_by(zip_code) %>%
  summarise(
    total = n(),
    vacant = sum(ovs),
    vacancy_rate = mean(ovs),
    .groups = "drop"
  ) %>%
  filter(total >= 100) %>%
  arrange(desc(vacancy_rate))

zip_summary_plot <- zip_summary %>%
  head(20) %>%
  ggplot(aes(x = reorder(as.character(zip_code), vacancy_rate), y = vacancy_rate)) +
  geom_col(fill = "tomato") +
  geom_text(aes(label = paste0(round(vacancy_rate * 100, 1), "%")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 20 Zip Codes by Vacancy Rate",
       subtitle = "Only zip codes with 100+ parcels",
       x = "Zip Code", y = "Vacancy Rate") +
  theme_minimal()

#ggsave("zip_summary.png", plot = zip_summary_plot, width = 10, height = 6, dpi = 300)

#```

```{r}
# Load necessary libraries
library(tidyverse)
library(tidycensus)
library(sf)
library(stringr)

# Note: You must have an active Census API key to use tidycensus
# census_api_key("YOUR_API_KEY", install = TRUE)

# 1. Fetch Philadelphia Census Tract geometries using tidycensus
# 1. Fetch Philadelphia Census Tract geometries
philly_tracts <- get_acs(
  geography = "tract",
  variables = "B01003_001", # Total Population
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  progress_bar = FALSE
) %>%
  select(GEOID, NAME, geometry)

# 2. Aggregate OVS data and calculate top 10
top_10_vacancy_tracts <- ovs_df %>%
  st_drop_geometry() %>%
  filter(!is.na(census_tract)) %>%
  mutate(
    # 1. Ensure it's numeric so we can do math
    tract_num = as.numeric(as.character(census_tract)),
    
    # 2. Multiply by 100 to handle the decimal suffix (e.g., 1 -> 100, 2.01 -> 201)
    tract_mult = tract_num * 100,
    
    # 3. Format as a 6-digit string with leading zeros
    clean_tract = sprintf("%06d", tract_mult),
    
    # 4. Attach the PA (42) and Philly (101) FIPS codes
    GEOID = paste0("42101", clean_tract)
  ) %>%
  group_by(GEOID) %>%
  summarise(
    total_parcels = n(),
    vacant_parcels = sum(ovs, na.rm = TRUE),
    vacancy_rate = vacant_parcels / total_parcels,
    .groups = "drop"
  ) %>%
  # Join back to get the Tract Name from the tidycensus object
  left_join(philly_tracts %>% st_drop_geometry(), by = "GEOID")

# 3. Display the result
top_10_vacancy_tracts %>%
  select(Tract_Name = GEOID, Total_Parcels = total_parcels, 
         Vacant_Parcels = vacant_parcels, Vacancy_Rate = vacancy_rate) %>%
  mutate(Vacancy_Rate = scales::percent(Vacancy_Rate, accuracy = 0.1)) %>%
  knitr::kable(caption = "Top 10 Philadelphia Census Tracts by Vacancy Rate")

# 4. Re-join the calculated data to the spatial geometries
philly_vacancy_map_data <- philly_tracts %>%
  left_join(
    # Selecting only necessary columns to avoid duplicating the 'NAME' column
    top_10_vacancy_tracts %>% select(GEOID, total_parcels, vacant_parcels, vacancy_rate), 
    by = "GEOID"
  )

# 5. Plot the Choropleth Map using ggplot2
ggplot(data = philly_vacancy_map_data) +
  # aes(fill = ...) tells ggplot to color the tracts based on the vacancy rate
  geom_sf(aes(fill = vacancy_rate), color = "white", linewidth = 0.01) +
  
  # A colorblind-friendly continuous color scale
  scale_fill_viridis_c(
    option = "inferno", # 'plasma' or 'viridis' are also great options here
    labels = scales::percent_format(accuracy = 1), 
    na.value = "gray90", # Colors tracts with NA (no data) light gray
    name = "Vacancy Rate"
  ) +
  
  # Labels and titles
  labs(
    title = "Philadelphia Vacancy Rates by Census Tract",
    subtitle = "Based on OVS Parcel Data",
    caption = "Gray areas indicate missing or filtered data."
  ) +
  
  # Clean up the map background and axes
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(b = 10)),
    legend.position = "right"
  )
```
```{r}
# 1. Align the coordinate systems, THEN spatially join
# First, transform ovs_df to match philly_tracts
ovs_df_projected <- st_transform(ovs_df, crs = st_crs(philly_tracts))

# Now, perform the spatial join with the matching CRSs
parcels_with_tracts <- st_join(
  ovs_df_projected, 
  philly_tracts %>% select(GEOID), 
  join = st_intersects
)

# 2. Aggregate the data (same as before)
tract_vacancy_rates <- parcels_with_tracts %>%
  st_drop_geometry() %>%
  filter(!is.na(GEOID)) %>% 
  group_by(GEOID) %>%
  summarise(
    total_parcels = n(),
    vacant_parcels = sum(ovs, na.rm = TRUE),
    vacancy_rate = vacant_parcels / total_parcels,
    .groups = "drop"
  )

# 3. Join back to tract geometries for the map (same as before)
philly_vacancy_map_data <- philly_tracts %>%
  left_join(tract_vacancy_rates, by = "GEOID")
```


```{r}
# 1. Create a spatial dataset of ONLY the vacant parcels
# Assuming 'ovs == 1' indicates vacancy; adjust if your column uses different values
vacant_parcels_sf <- ovs_df %>% 
  filter(ovs == 1) 

# 2. Plot the layered map
ggplot() +
  # BOTTOM LAYER: The Census Tract Choropleth
  geom_sf(
    data = philly_vacancy_map_data, 
    aes(fill = vacancy_rate), 
    color = "white", 
    linewidth = 0.1
  ) +
  
  # TOP LAYER: The Vacant Parcels Overlay
  geom_sf(
    data = vacant_parcels_sf, 
    color = "black",        # Color of the parcels/points
    fill = "black",         # Fill color if they are polygons
    size = 0.1,             # Make them small so they don't overwhelm the map
    alpha = 0.3             # Make them transparent so the tract colors show through
  ) +
  
  # A colorblind-friendly continuous color scale for the base map
  scale_fill_viridis_c(
    option = "inferno", 
    labels = scales::percent_format(accuracy = 1), 
    na.value = "gray90", 
    name = "Vacancy Rate"
  ) +
  
  # Labels and titles
  labs(
    title = "Philadelphia Vacancy Rates with Parcel Overlay",
    subtitle = "Black dots/polygons represent individual vacant parcels",
    caption = "Data: OVS and US Census Bureau"
  ) +
  
  # Clean up the map background and axes
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(b = 10)),
    legend.position = "right"
  )
```

```{r}
# Load necessary libraries
library(tidyverse)
library(tidycensus)
library(sf)

# Note: You must have an active Census API key to use tidycensus
# census_api_key("YOUR_API_KEY", install = TRUE)

# 1. Fetch Philadelphia Census Tract geometries using tidycensus
philly_tracts <- get_acs(
  geography = "tract",
  variables = "B01003_001", # Total Population
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  progress_bar = FALSE
) %>%
  select(GEOID, NAME, geometry)

# 2. Align CRS and Spatially Join Parcels to Tracts
# Transform ovs_df to match the coordinate reference system of philly_tracts
ovs_df_projected <- st_transform(ovs_df, crs = st_crs(philly_tracts))

# Perform the spatial join to assign a GEOID to every parcel based on location
parcels_with_tracts <- st_join(
  ovs_df_projected, 
  philly_tracts %>% select(GEOID), 
  join = st_intersects
)

# 3. Aggregate Vacancy Data for ALL Tracts
tract_vacancy_rates <- parcels_with_tracts %>%
  st_drop_geometry() %>%
  filter(!is.na(GEOID)) %>% 
  group_by(GEOID) %>%
  summarise(
    total_parcels = n(),
    vacant_parcels = sum(ovs, na.rm = TRUE),
    vacancy_rate = vacant_parcels / total_parcels,
    .groups = "drop"
  )

# 4. Join Calculated Rates back to Tract Geometries for Mapping
philly_vacancy_map_data <- philly_tracts %>%
  left_join(tract_vacancy_rates, by = "GEOID")

# 5. Create a spatial dataset of ONLY the vacant parcels for the overlay
# Using the projected dataframe ensures it maps perfectly over the tracts
vacant_parcels_sf <- ovs_df_projected %>% 
  filter(ovs == 1) 

# 6. Plot the Layered Map using ggplot2
vacant_tract <- ggplot() +
  # BOTTOM LAYER: The Census Tract Choropleth (All Tracts)
  geom_sf(
    data = philly_vacancy_map_data, 
    aes(fill = vacancy_rate), 
    color = "white", 
    linewidth = 0.01
  ) +
  
  # TOP LAYER: The Vacant Parcels Overlay
 # geom_sf(
  #  data = vacant_parcels_sf, 
  # color = "red",        
  #  fill = "red",         
   # size = 0.1,             
  #  alpha = 0.1             
  #) +
  
  # A colorblind-friendly continuous color scale
  scale_fill_viridis_c(
    option = "mako", 
    direction = -1,
    labels = scales::percent_format(accuracy = 1), 
    na.value = "gray90", 
    name = "Vacancy Rate"
  ) +
  
  # Labels and titles
  labs(
    title = "Philadelphia Vacancy Rates by Census Tract",
    caption = "Data: OVS and US Census Bureau | Gray tracts indicate missing data"
  ) +
  
  # Clean up the map background and axes
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(b = 10)),
    legend.position = "right"
  )
ggsave("census.png", plot = vacant_tract, width = 10, height = 10)
```

