---
title: "PhillyStat360 — 06: Vector Tiling with Tippecanoe (R port)"
output: html_document
---

R port of `code/python/06_tiling.ipynb`. Builds vector tilesets (`.pmtiles`)
from the GeoJSON exports produced by `05_output_analysis.Rmd`.

> Tippecanoe is a CLI tool — there is no R wrapper. This Rmd shells out to
> `tippecanoe` via `system2()`. Install it first:
>
> - Linux/Colab: `apt-get install -y tippecanoe`
> - macOS: `brew install tippecanoe`
> - Windows: build the [Felt fork](https://github.com/felt/tippecanoe) under WSL or use Docker
>
> The tippecanoe binary must be on `PATH` for the chunks below to run.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 0. Setup

```{r paths}
ROOT      <- here::here()
PY_PATH   <- file.path(ROOT, "data_py")
TILE_PATH <- file.path(PY_PATH, "tiles")
dir.create(TILE_PATH, recursive = TRUE, showWarnings = FALSE)

FULL_GEOJSON    <- file.path(PY_PATH, "vacancy_predictions.geojson")
FLAGGED_GEOJSON <- file.path(PY_PATH, "vacancy_predictions_flagged.geojson")

for (p in c(FULL_GEOJSON, FLAGGED_GEOJSON)) {
  if (!file.exists(p)) {
    message("[!] missing: ", p, "  — run 05_output_analysis.Rmd §6 first")
  } else {
    cat(sprintf("[ok] %s: %.1f MB\n",
                basename(p), file.info(p)$size / 1e6))
  }
}
```

## 1. Verify tippecanoe install

```{r tippecanoe-check}
have_tippecanoe <- function() nzchar(Sys.which("tippecanoe"))
stopifnot(have_tippecanoe())
system2("tippecanoe", "--version")
```

## 2. Tile the flagged-only set (small, fast)

```{r tile-flagged}
flagged_pmtiles <- file.path(TILE_PATH, "vacancy_flagged.pmtiles")
args <- c(
  "-o", shQuote(flagged_pmtiles),
  "-l", "flagged",
  "-n", shQuote("PhillyStat360 Flagged Parcels (top 1%)"),
  "-Z10", "-z16",
  "--coalesce-densest-as-needed",
  "--extend-zooms-if-still-dropping",
  "--force",
  shQuote(FLAGGED_GEOJSON)
)
cat("tippecanoe", paste(args, collapse = " "), "\n")
status <- system2("tippecanoe", args)
stopifnot(status == 0)
cat(sprintf("Wrote %s (%.1f MB)\n",
            flagged_pmtiles, file.info(flagged_pmtiles)$size / 1e6))
```

## 3. Tile the full prediction set

```{r tile-full}
full_pmtiles <- file.path(TILE_PATH, "vacancy_predictions.pmtiles")
args <- c(
  "-o", shQuote(full_pmtiles),
  "-l", "parcels",
  "-n", shQuote("PhillyStat360 Vacancy Predictions"),
  "-Z10", "-z15",
  "--drop-densest-as-needed",
  "--extend-zooms-if-still-dropping",
  "--simplification=10",
  "--force",
  shQuote(FULL_GEOJSON)
)
cat("tippecanoe", paste(args, collapse = " "), "\n")
status <- system2("tippecanoe", args)
stopifnot(status == 0)
cat(sprintf("Wrote %s (%.1f MB)\n",
            full_pmtiles, file.info(full_pmtiles)$size / 1e6))
```

## 4. Verify outputs

```{r verify}
for (pmt in c(flagged_pmtiles, full_pmtiles)) {
  cat("===", basename(pmt), "===\n")
  cat(sprintf("Size: %.1f MB\n", file.info(pmt)$size / 1e6))
  # First metadata line via tippecanoe-decode
  tryCatch({
    out <- system2("tippecanoe-decode",
                   c("-s", "EPSG:4326", "-c", shQuote(pmt)),
                   stdout = TRUE, stderr = FALSE)
    cat(out[1], "\n\n")
  }, error = function(e) message(e$message))
}
```
