---
title: "PhillyStat360 — 03a: OVS Construction"
author: "MUSA Practicum Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(sf)
library(lubridate)
library(broom)
library(pROC)
library(caret)

data_path  <- "../rawdata/"
out_path   <- "../data/"
graph_path <- "../graphs/"

dir.create(out_path,   showWarnings = FALSE, recursive = TRUE)
dir.create(graph_path, showWarnings = FALSE, recursive = TRUE)

# Temporal Anchor: All predictive variables must strictly precede this date
TRAIN_CUTOFF  <- as.Date("2025-10-01")
RECENT_WINDOW <- TRAIN_CUTOFF - months(6)
```

---

## 1. load data

```{r read-data, cache=TRUE}
properties <- read_csv(paste0(data_path, "opa_properties_public.csv"),
                       show_col_types = FALSE) %>% clean_names()
clean_seal  <- read_csv(paste0(data_path, "clean_seal.csv"),
                        show_col_types = FALSE) %>% clean_names()
violations  <- read_csv(paste0(data_path, "violations.csv"),
                        show_col_types = FALSE) %>% clean_names()
licenses    <- read_csv(paste0(data_path, "business_licenses.csv"),
                        show_col_types = FALSE) %>% clean_names()
unsafe      <- read_csv(paste0(data_path, "unsafe.csv"),
                        show_col_types = FALSE) %>% clean_names()
imm_dang    <- read_csv(paste0(data_path, "imm_dang.csv"),
                        show_col_types = FALSE) %>% clean_names()
parcel_sf   <- st_read(paste0(data_path, "PWD_PARCELS.geojson"),
                       quiet = TRUE) %>% clean_names()
permits    <- read_csv(paste0(data_path, "permits_2020.csv"),
                       show_col_types = FALSE) %>% clean_names()

# Load VIOLATION_DEFINITION.csv if available for tier mapping display
violation_def_path <- paste0(data_path, "VIOLATION_DEFINITION.csv")
if (file.exists(violation_def_path)) {
  violation_def <- read_csv(violation_def_path, show_col_types = FALSE) %>%
    clean_names()
} else {
  violation_def <- NULL
  message("VIOLATION_DEFINITION.csv not found; skipping code title display.")
}

cat("properties:", comma(nrow(properties)), "\n")
cat("clean_seal:", comma(nrow(clean_seal)), "\n")
cat("violations:", comma(nrow(violations)), "\n")
cat("licenses:  ", comma(nrow(licenses)),   "\n")
cat("unsafe:    ", comma(nrow(unsafe)),     "\n")
cat("permits:   ", comma(nrow(permits)),    "\n")
cat("imm_dang:  ", comma(nrow(imm_dang)),   "\n")
cat("parcel_sf: ", comma(nrow(parcel_sf)),  "\n")
```

---

## 2. Construct OVS (Observed Vacancy Status)

Following the VacancyGuide2026 methodology, OVS=1 if any of the following are met:

- **Source 1: Clean & Seal** — Property appears in L&I Clean & Seal records
- **Source 2: Vacant Violations** — Has vacancy-related violation codes with status = 'OPEN'.
- **Source 3: Vacant License** — Holds an active vacant property license (Residential 3219 or Commercial 3634).

### 2a. Source 1 — Clean & Seal

```{r ovs-cleanseal}
# Identify parcels with completed demolition permits following a Clean & Seal event
demo_after_seal <- clean_seal %>%
  filter(!is.na(opa_account_num)) %>%
  select(opa_account_num, casecreateddate) %>%
  inner_join(
    permits %>%
      filter(tolower(permittype) == "demolition",
             !is.na(permitcompleteddate),
             tolower(status) == "completed") %>%
      select(opa_account_num, permitcompleteddate),
    by = "opa_account_num"
  ) %>%
  filter(permitcompleteddate > casecreateddate) %>%
  distinct(opa_account_num)

cs_parcels <- clean_seal %>%
  filter(!is.na(opa_account_num)) %>%
  # Fichman fix 1: within a 2 year window
  filter(casecreateddate >= (TRAIN_CUTOFF - lubridate::years(2))) %>%
  # Fichman fix 2: demolition after clean and seal = empty lot
  anti_join(demo_after_seal, by = "opa_account_num") %>%
  distinct(opa_account_num) %>%
  mutate(ovs_clean_seal = 1)

cat("Unique Clean & Seal parcels (2yr window + demolition filter):", comma(nrow(cs_parcels)), "\n")
```

### 2b. Source 2 — Open Vacant Violations

```{r ovs-violations}
# 11 vacancy violation codes used for OVS definition (not used as predictors)
ovs_viol_codes <- c(
  "9-3904", "9-3905", "PM15-901.1", "PM15-901.2", "PM15-301",
  "CP-102",  "CP-103", "PM-102.4/1", "PM-102.4/2",
  "PM-102.4/3", "PM-102.4/4"
)

viol_parcels <- violations %>%
  filter(violationcode %in% ovs_viol_codes,
         tolower(violationstatus) == "open") %>%
  distinct(opa_account_num) %>%
  filter(!is.na(opa_account_num)) %>%
  mutate(ovs_violation = 1)

cat("Parcels with OPEN vacant violations:", comma(nrow(viol_parcels)), "\n")

violations %>%
  filter(violationcode %in% ovs_viol_codes,
         tolower(violationstatus) == "open") %>%
  count(violationcode, sort = TRUE) %>%
  kable(caption = "Distribution of OPEN Vacant Violation Codes") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### 2c. Source 3 — Active Vacant License

```{r ovs-license}
# 3219 = Residential Vacant, 3634 = Commercial Vacant
vacant_license_parcels <- licenses %>%
  filter(tolower(licensestatus) == "active",
         revenuecode %in% c("3219", "3634")) %>%
  distinct(opa_account_num) %>%
  filter(!is.na(opa_account_num)) %>%
  mutate(ovs_license = 1)

cat("Parcels with active vacant licenses:", comma(nrow(vacant_license_parcels)), "\n")
```

### 2d. Merge OVS to Properties

```{r ovs-merge}
ovs_df <- properties %>%
  select(parcel_number, location, owner_1, category_code, category_code_description,
         year_built, total_livable_area, market_value, sale_date, sale_price,
         exterior_condition, interior_condition,
         zoning, zip_code, census_tract) %>%
  left_join(cs_parcels,             by = c("parcel_number" = "opa_account_num")) %>%
  left_join(viol_parcels,           by = c("parcel_number" = "opa_account_num")) %>%
  left_join(vacant_license_parcels, by = c("parcel_number" = "opa_account_num")) %>%
  replace_na(list(ovs_clean_seal = 0, ovs_violation = 0, ovs_license = 0)) %>%
  mutate(
    ovs         = if_else(ovs_clean_seal == 1 | ovs_violation == 1 | ovs_license == 1, 1, 0),
    ovs_sources = ovs_clean_seal + ovs_violation + ovs_license
  )

cat("Total parcels:", comma(nrow(ovs_df)), "\n")
cat("OVS=1:  ", comma(sum(ovs_df$ovs == 1)),
    "(", percent(mean(ovs_df$ovs), 0.1), ")\n")
cat("OVS=0:  ", comma(sum(ovs_df$ovs == 0)), "\n")
```

### 2e. Filter for Residential + Join Geometry

Includes: Single Family (1), Multi-Family (2), Mixed (3), Large Apartments > 4 units (14).

```{r filter-attach}
ovs_df <- ovs_df %>%
  filter(category_code %in% c("1", "2", "3", "14")) %>%
  left_join(
    parcel_sf %>% select(brt_id, bldg_desc, shape_area, shape_length, geometry),
    by = c("parcel_number" = "brt_id")
  ) %>%
  st_as_sf()

cat("Post-filter parcels:", comma(nrow(ovs_df)), "\n")
cat("OVS=1:", comma(sum(ovs_df$ovs == 1)),
    "(", percent(mean(ovs_df$ovs), 0.1), ")\n")
```

### 2f. OVS Source Overlap Analysis

```{r ovs-overlap}
ovs_positive <- ovs_df %>% filter(ovs == 1)

cat("=== OVS=1 Source Breakdown ===\n")
cat("Clean & Seal Only:", comma(sum(ovs_positive$ovs_clean_seal==1 & ovs_positive$ovs_violation==0 & ovs_positive$ovs_license==0)), "\n")
cat("Violation Only:   ", comma(sum(ovs_positive$ovs_clean_seal==0 & ovs_positive$ovs_violation==1 & ovs_positive$ovs_license==0)), "\n")
cat("License Only:     ", comma(sum(ovs_positive$ovs_clean_seal==0 & ovs_positive$ovs_violation==0 & ovs_positive$ovs_license==1)), "\n")
cat("Two-source overlap:       ", comma(sum(ovs_positive$ovs_sources == 2)), "\n")
cat("Three-source overlap:       ", comma(sum(ovs_positive$ovs_sources == 3)), "\n")
```

```{r ovs-overlap-plot, fig.width=8, fig.height=5}
p_combo <- ovs_positive %>%
  st_drop_geometry() %>%
  mutate(source_combo = case_when(
    ovs_clean_seal==1 & ovs_violation==1 & ovs_license==1 ~ "All Three",
    ovs_clean_seal==1 & ovs_violation==1 ~ "Clean&Seal + Violation",
    ovs_clean_seal==1 & ovs_license==1   ~ "Clean&Seal + License",
    ovs_violation==1  & ovs_license==1   ~ "Violation + License",
    ovs_clean_seal==1 ~ "Clean & Seal Only",
    ovs_violation==1  ~ "Violation Only",
    ovs_license==1    ~ "License Only"
  )) %>%
  count(source_combo, sort = TRUE) %>%
  ggplot(aes(x = reorder(source_combo, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Distribution of OVS=1 Parcels by Source Combination", x = NULL, y = "parcel count") +
  theme_minimal()

p_combo
ggsave(paste0(graph_path, "ovs_combination_plot.png"),
       plot = p_combo, width = 10, height = 6, dpi = 300)
```

---

## 3. Export

```{r export-ovs}
# Drop geometry for downstream flat-file use in 03_analysis and 03_features
ovs_df %>%
  st_drop_geometry() %>%
  write_csv(paste0(out_path, "ovs_residential.csv"))

cat("Exported: ovs_residential.csv —", comma(nrow(ovs_df)), "rows\n")
cat("OVS=1:", comma(sum(ovs_df$ovs == 1)),
    "(", percent(mean(ovs_df$ovs), 0.1), ")\n")
cat("Columns:", ncol(ovs_df) - 1, "(geometry dropped)\n")
```

---
