---
title: "PhillyStat360 — 03b: Tier Mapping, EDA & Baseline Model"
author: "MUSA Practicum Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(sf)
library(lubridate)
library(broom)
library(pROC)
library(caret)

data_path  <- "../rawdata/"
out_path   <- "../data/"
graph_path <- "../graphs/"

dir.create(out_path,   showWarnings = FALSE, recursive = TRUE)
dir.create(graph_path, showWarnings = FALSE, recursive = TRUE)

TRAIN_CUTOFF  <- as.Date("2025-10-01")
RECENT_WINDOW <- TRAIN_CUTOFF - months(6)
```

---

## 1. Load Data

```{r load-data, cache=TRUE}
# OVS output from 03_ovs.Rmd
ovs_df <- read_csv(paste0(out_path, "ovs_residential.csv"),
                   show_col_types = FALSE)

# Raw violations needed for Section 3c tier validation
violations <- read_csv(paste0(data_path, "violations.csv"),
                       show_col_types = FALSE) %>% clean_names()

# Violation definitions (optional, for tier mapping display)
violation_def_path <- paste0(data_path, "VIOLATION_DEFINITION.csv")
if (file.exists(violation_def_path)) {
  violation_def <- read_csv(violation_def_path, show_col_types = FALSE) %>%
    clean_names()
} else {
  violation_def <- NULL
  message("VIOLATION_DEFINITION.csv not found; skipping code title display.")
}

# OVS violation codes (re-defined here for use in Section 3b in_ovs_def flag)
ovs_viol_codes <- c(
  "9-3904", "9-3905", "PM15-901.1", "PM15-901.2", "PM15-301",
  "CP-102",  "CP-103", "PM-102.4/1", "PM-102.4/2",
  "PM-102.4/3", "PM-102.4/4"
)

cat("ovs_df:", comma(nrow(ovs_df)), "parcels |",
    "OVS=1:", percent(mean(ovs_df$ovs), 0.1), "\n")
cat("violations:", comma(nrow(violations)), "\n")
```

---

## 3. VacancyGuide Violation Tier Mapping

Constructs the violation code tier table (108 codes) for feature engineering.

VacancyGuide2026 has three tiers（+1/+2/+3), totalling 108 codes。

**note（Fichman feedback, 2026-02-18）：**
Tier weights were established via trial-and-error without statistical validation. （Tim Haynes at 1/27 kickoff meeting). 

We use these tiers for for Track B comparison rather than primary predictors.

### 3a. Define VacancyGuide Code Lists

```{r vacancyguide-codes}
# Level 1: +1 Point — Minor deterioration and general maintenance
guide_level1 <- c(
  "PM15-304.10", "PM15-304.11", "PM15-304.19", "PM15-304.19O",
  "PM15-304.2",  "PM15-304.8",  "PM15-308.1",  "PM15-309.1",
  "PM15-309.3",  "PM15-309.4",  "PM-302.2/1",  "PM-302.2/2",
  "PM-302.3/1",  "PM-302.4/6",  "PM-302.4/7",  "PM-304.1/1",
  "PM-304.6/1",  "PM-304.8/1",  "PM-304.8/15", "PM-304.8/16",
  "PM-304.8/17", "PM-304.8/18", "PM-304.8/19", "PM-304.8/2",
  "PM-304.8/20", "PM-304.8/21"
)

# Level 2: +2 Points — Moderate structural damage
guide_level2 <- c(
  "PM15-110.7",  "PM15-302.1",  "PM15-302.4",  "PM15-302.5",
  "PM15-304.13", "PM15-304.15", "PM15-304.16", "PM15-304.4",
  "PM15-304.5",  "PM15-304.6",  "PM15-304.7",  "PM-304.2/1",
  "PM-304.3/1",  "PM-304.3/2",  "PM-304.3/3",  "PM-304.3/4",
  "PM-304.3/5",  "PM-304.5/2",  "PM-304.5/3"
)

# Level 3: +3 Points — Severe: Vacant/Unsafe/Collapse
guide_level3 <- c(
  "9-3905",       "CP-103",       "PM-102.4/2",   "PM-102.4/4",
  "PM15-108.1",   "PM15-108.2",   "PM15-109.1",   "PM15-110.1",
  "PM15-301",     "PM15-304.1",   "PM15-304.19V", "PM15-304.1A",
  "PM15-304.1B",  "PM15-304.1C",  "PM15-304.1D",  "PM15-304.1E",
  "PM15-304.1F",  "PM15-304.1G",  "PM15-304.1H",  "PM15-304.1I",
  "PM15-304.1J",  "PM15-901.1",   "PM15-901.2",   "PM15-902.1",
  "PM-304.4/1",   "PM-306.0/1",   "PM-306.0/90",  "PM-306.2/1",
  "PM-306.2/4",   "PM-306.2/90",  "PM-306.2/91",  "PM-306.2/92",
  "PM-307.1/1",   "PM-307.1/11",  "PM-307.1/12",  "PM-307.1/13",
  "PM-307.1/19",  "PM-307.1/2",   "PM-307.1/3",   "PM-307.1/4",
  "PM-307.1/5",   "PM-307.1/6",   "PM-307.1/7",   "PM-307.1/8",
  "PM-307.1/9",   "PM-307.1/90",  "PM-307.2/1",   "PM-307.3/1",
  "PM-307.3/2",   "PM-307.3/90",  "PM-307.3/91",  "PM-308.1/1",
  "PM-308.1/10",  "PM-308.1/11",  "PM-308.1/12",  "PM-308.1/2",
  "PM-308.1/3",   "PM-308.1/4",   "PM-308.1/5",   "PM-308.1/6",
  "PM-308.1/7",   "PM-308.1/8",   "PM-308.1/9"
)

# Vacant Land Codes (For reference only; building model does not use these)
guide_land <- c(
  "CP-102", "PM-102.4/1", "PM-102.4/3", "PM-302.2/4",
  "PM-304.9/2", "PM-304.9/3", "PM-304.9/4", "PM-306.0/2", "PM-306.0/91"
)

building_all <- c(guide_level1, guide_level2, guide_level3)

cat("Level 1 (+1分):", length(guide_level1), "codes\n")
cat("Level 2 (+2分):", length(guide_level2), "codes\n")
cat("Level 3 (+3分):", length(guide_level3), "codes\n")
cat("Building Model Total: ", length(building_all), "codes\n")
cat("Land Model:     ", length(guide_land),   "codes\n")

# Duplicate verification
stopifnot("Duplicate codes found across building tiers!" =
          length(building_all) == length(unique(building_all)))
cat("No duplicate codes between building tiers\n")
```

### 3b. Construct Full Tier Mapping Table

```{r complete-mapping}
building_tier_map <- bind_rows(
  tibble(code = guide_level1, guide_tier = "Level 1", guide_weight = 1L),
  tibble(code = guide_level2, guide_tier = "Level 2", guide_weight = 2L),
  tibble(code = guide_level3, guide_tier = "Level 3", guide_weight = 3L)
) %>%
  mutate(in_ovs_def = code %in% ovs_viol_codes)

# Append titles if violation_def exists
if (!is.null(violation_def)) {
  building_tier_map <- building_tier_map %>%
    left_join(violation_def %>% select(violation_code, violation_code_title),
              by = c("code" = "violation_code"))
}

building_tier_map %>%
  count(guide_tier, guide_weight) %>%
  kable(caption = "VacancyGuide Building Violation Tier Summary") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### 3c. Validation: Tier Code Frequency in Real Data

```{r validate-tiers}
building_viol_counts <- violations %>%
  filter(violationcode %in% building_all,
         tolower(violationstatus) == "open") %>%
  count(violationcode, name = "n_open") %>%
  right_join(building_tier_map %>% select(code, guide_tier, guide_weight),
             by = c("violationcode" = "code")) %>%
  replace_na(list(n_open = 0)) %>%
  arrange(guide_weight, desc(n_open))

building_viol_counts %>%
  group_by(guide_tier) %>%
  summarise(
    codes_with_data = sum(n_open > 0),
    total_codes     = n(),
    total_open      = sum(n_open),
    .groups = "drop"
  ) %>%
  kable(col.names = c("Tier","Codes w/ Data", "Total Codes", "Total OPEN Violations"),
        caption = "OPEN Violation Counts per Tier in Current Data") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### 3d. 54 Vacancy Keyword Codes (Track A)

```{r vacancy-keyword-codes}
# Screen for codes where title or definition contains "vacant"
if (!is.null(violation_def)) {
  vacancy_keyword_codes <- violation_def %>%
    filter(str_detect(tolower(violation_code_title), "vacant") |
           str_detect(tolower(violation_definition),  "vacant")) %>%
    pull(violation_code)
  cat("Vacancy keyword codes found:", length(vacancy_keyword_codes), "\n")
} else {
  # fallback: use known vacancy codes from Level 3
  vacancy_keyword_codes <- c(
    ovs_viol_codes,
    "PM15-108.1", "PM15-108.2", "PM15-109.1", "PM15-110.1",
    "PM15-304.1",  "PM15-304.19V", "PM15-901.1", "PM15-901.2",
    "PM15-902.1",  "PM-306.0/1",   "PM-306.2/1", "PM-307.1/1",
    "PM-307.2/1",  "PM-307.3/1"
  ) %>% unique()
  cat("Using fallback vacancy codes, count:", length(vacancy_keyword_codes), "\n")
}

# Maintenance codes (Level 1+2 excluding vacancy keywords)
maintenance_codes <- building_tier_map %>%
  filter(guide_weight %in% c(1, 2)) %>%
  pull(code)

# Structural codes (Level 3)
structural_codes <- building_tier_map %>%
  filter(guide_weight == 3) %>%
  pull(code)
```

---

## 4. EDA — Key Variable Exploration

### 4a. Exterior Condition & Vacancy Rate

> Exterior condition 是 1–7 的有序分类变量，使用 bar chart 展示各等级空置率。
> （Fichman feedback：violin plot 不适合有序分类变量）

```{r eda-exterior, fig.width=7, fig.height=4}
p_ext <- ovs_df %>%
  filter(!is.na(exterior_condition)) %>%
  mutate(exterior_condition = factor(as.integer(exterior_condition))) %>%
  group_by(exterior_condition) %>%
  summarise(n = n(), n_vacant = sum(ovs == 1),
            vac_rate = mean(ovs == 1), .groups = "drop") %>%
  ggplot(aes(x = exterior_condition, y = vac_rate, fill = exterior_condition)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(vac_rate, 0.1)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "OrRd") +
  labs(title    = "Vacancy Rate by Exterior Condition Rating",
       subtitle = "1 = New/Excellent → 7 = Collapsed/Deteriorated",
       x = "Exterior Condition (1–7)", y = "OVS=1 Rate") +
  theme_minimal()

p_ext
ggsave(paste0(graph_path, "exterior_condition_vacancy_rate.png"),
       plot = p_ext, width = 8, height = 5, dpi = 300)
```

### 4b. OVS=1 Analysis in New Construction

> Fichman feedback：为什么新建（post-2010）房产会出现在 OVS=1 中？

```{r eda-new-vacant}
new_vacant <- ovs_df %>%
  filter(!is.na(year_built), as.integer(year_built) >= 2010, ovs == 1)

cat("Parcels built after 2010 with OVS=1:", nrow(new_vacant), "\n\n")

bind_rows(
  new_vacant %>%
    summarise(group = "New Construction (2010+) & Vacant",
              pct_clean_seal = mean(ovs_clean_seal == 1),
              pct_violation  = mean(ovs_violation  == 1),
              pct_license    = mean(ovs_license    == 1)),
  ovs_df %>% filter(ovs == 1) %>%
    summarise(group = "All OVS=1 (Baseline)",
              pct_clean_seal = mean(ovs_clean_seal == 1),
              pct_violation  = mean(ovs_violation  == 1),
              pct_license    = mean(ovs_license    == 1))
) %>%
  mutate(across(where(is.numeric), ~percent(.x, 0.1))) %>%
  kable(col.names = c("Group", "Clean & Seal %", "Violation %", "License %"),
        caption = "New Vacant vs. All Vacant: OVS Source Composition") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### 4c. Violation History vs. Vacancy (Fichman: faceted plot)

> Fichman feedback：画出过去不同时间窗口内违规数量均值，按 OVS=1/0 分组对比。
> 这是验证 life history features 有意义的核心 EDA。

```{r eda-violations-faceted, fig.width=10, fig.height=6}
# Violation counts in different windows (all codes included for frequency analysis)
viol_for_eda <- violations %>%
  filter(!is.na(opa_account_num)) %>%
  mutate(
    violationdate = as.Date(violationdate),
    is_pre        = !is.na(violationdate) & violationdate < TRAIN_CUTOFF,
    is_5yr        = !is.na(violationdate) &
                      violationdate >= (TRAIN_CUTOFF - lubridate::years(5)) &
                      violationdate < TRAIN_CUTOFF,
    is_3yr        = !is.na(violationdate) &
                      violationdate >= (TRAIN_CUTOFF - lubridate::years(3)) &
                      violationdate < TRAIN_CUTOFF,
    is_recent     = !is.na(violationdate) &
                      violationdate >= RECENT_WINDOW & violationdate < TRAIN_CUTOFF
  ) %>%
  filter(is_pre) %>%
  group_by(opa_account_num) %>%
  summarise(
    n_viol_total  = n(),
    n_viol_5yr    = sum(is_5yr),
    n_viol_3yr    = sum(is_3yr),
    n_viol_recent = sum(is_recent),
    .groups = "drop"
  )

viol_eda_df <- ovs_df %>%
  select(parcel_number, ovs) %>%
  left_join(viol_for_eda, by = c("parcel_number" = "opa_account_num")) %>%
  mutate(
    across(starts_with("n_viol_"), ~replace_na(.x, 0)),
    ovs_label = if_else(ovs == 1, "OVS=1 (Vacant)", "OVS=0 (Occupied)")
  )

# 按组计算均值，转长格式
viol_means <- viol_eda_df %>%
  group_by(ovs_label) %>%
  summarise(
    `All Time`      = mean(n_viol_total),
    `Last 5 Years`  = mean(n_viol_5yr),
    `Last 3 Years`  = mean(n_viol_3yr),
    `Last 6 Months` = mean(n_viol_recent),
    .groups = "drop"
  ) %>%
  pivot_longer(-ovs_label, names_to = "window", values_to = "mean_violations") %>%
  mutate(window = factor(window,
                         levels = c("All Time", "Last 5 Years",
                                    "Last 3 Years", "Last 6 Months")))

p_viol_facet <- viol_means %>%
  ggplot(aes(x = ovs_label, y = mean_violations, fill = ovs_label)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = round(mean_violations, 2)),
            vjust = -0.4, size = 3.5) +
  facet_wrap(~window, scales = "free_y") +
  scale_fill_manual(values = c("OVS=0 (Occupied)" = "steelblue",
                                "OVS=1 (Vacant)"  = "tomato")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title    = "Mean Number of Violations by Vacancy Status",
       subtitle = "Across different time windows | properties with no violations counted as 0",
       x = NULL, y = "Mean Violations per Parcel") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))

p_viol_facet
ggsave(paste0(graph_path, "violations_faceted_by_ovs.png"),
       plot = p_viol_facet, width = 10, height = 6, dpi = 300)
```

---

## 5. Baseline Model (OPA Fields Only)

Baseline 仅使用已在 `ovs_df` 上的 OPA 字段，不做任何额外 join。
这是模型性能的下限，03_features.Rmd 的 engineered model 必须超越此数字。

| 纳入 | 理由 |
|------|------|
| `exterior_condition` | 评估员直接观测的物理状态 |
| `log(market_value)` | 经济去投资信号，对数处理消除右偏 |
| `building_age` | 由 `year_built` 推算 |
| `log(total_livable_area)` | 建筑面积代理 |

### 5a. Prepare Dataset

```{r baseline-prep}
baseline_df <- ovs_df %>%
  filter(market_value > 0, total_livable_area > 0, !is.na(exterior_condition)) %>%
  mutate(
    exterior_condition = as.numeric(exterior_condition),
    building_age       = 2024L - as.integer(year_built),
    building_age       = if_else(building_age < 0 | building_age > 200,
                                 NA_integer_, building_age),
    log_market_value   = log(market_value),
    log_livable_area   = log(total_livable_area)
  ) %>%
  filter(!is.na(building_age))

cat("Baseline 数据集:", comma(nrow(baseline_df)), "地块\n")
cat("OVS=1:", comma(sum(baseline_df$ovs == 1)),
    "(", percent(mean(baseline_df$ovs), 0.1), ")\n")
```

### 5b. Fit model

```{r baseline-model}
baseline_model <- glm(
  ovs ~ exterior_condition + log_market_value + building_age + log_livable_area,
  data = baseline_df, family = binomial()
)
summary(baseline_model)
```

### 5c. Fit Statistics & Odds Ratios

```{r baseline-fit}
null_dev    <- baseline_model$null.deviance
resid_dev   <- baseline_model$deviance

tibble(
  Metric = c("McFadden R²", "AUC (训练集)", "AIC", "N"),
  Value  = c(round(1 - resid_dev/null_dev, 3),
             "—（见下方 ROC）",
             round(AIC(baseline_model), 1),
             comma(nobs(baseline_model)))
) %>%
  kable(caption = "Baseline 拟合统计") %>%
  kable_styling(full_width = FALSE)

tidy(baseline_model, exponentiate = TRUE, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~round(.x, 3)),
         p.value = format.pval(p.value, digits = 3)) %>%
  kable(caption = "Baseline Odds Ratios") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

### 5d. Threshold & Probability Distribution

> Fichman feedback：用预测概率密度图代替单一阈值数字讨论。

```{r baseline-threshold}
pred_probs_baseline <- predict(baseline_model, type = "response")
actual_baseline     <- baseline_df$ovs

roc_baseline  <- roc(actual_baseline, pred_probs_baseline, quiet = TRUE)
best_baseline <- coords(roc_baseline, "best", best.method = "youden",
                        ret = c("threshold", "sensitivity", "specificity"))

cat("Baseline AUC:       ", round(auc(roc_baseline), 3), "\n")
cat("Youden 最优阈值:    ", round(best_baseline$threshold, 3), "\n")
cat("Sensitivity:        ", round(best_baseline$sensitivity, 3), "\n")
cat("Specificity:        ", round(best_baseline$specificity, 3), "\n")
```

```{r baseline-prob-dist, fig.width=8, fig.height=5}
p_prob_baseline <- tibble(
  prob  = pred_probs_baseline,
  label = factor(actual_baseline, levels = c(0,1),
                 labels = c("OVS=0 (Occupied)", "OVS=1 (Vacant)"))
) %>%
  ggplot(aes(x = prob, fill = label)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = best_baseline$threshold,
             linetype = "dashed", color = "black", linewidth = 0.8) +
  annotate("text", x = best_baseline$threshold + 0.005, y = Inf,
           label = paste0("Youden = ", round(best_baseline$threshold, 3)),
           vjust = 2, hjust = 0, size = 3.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_x_continuous(labels = percent_format()) +
  labs(title    = "Baseline: Predicted Probability Distribution by True OVS",
       subtitle = paste0("AUC = ", round(auc(roc_baseline), 3),
                         " | 重叠多 = 分离差；两侧分开 = 分离好"),
       x = "Predicted P(Vacant)", y = "Density", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p_prob_baseline
ggsave(paste0(graph_path, "baseline_prob_distribution.png"),
       plot = p_prob_baseline, width = 9, height = 5, dpi = 300)
```

```{r baseline-roc, fig.width=6, fig.height=5}
ggroc(roc_baseline, color = "steelblue", linewidth = 1) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray50") +
  annotate("point", x = best_baseline$specificity,
           y = best_baseline$sensitivity, color = "tomato", size = 4) +
  labs(title    = "ROC Curve — Baseline Model",
       subtitle = paste0("AUC = ", round(auc(roc_baseline), 3)),
       x = "Specificity", y = "Sensitivity") +
  theme_minimal()
```

```{r baseline-cm}
pred_class_baseline <- if_else(pred_probs_baseline >= best_baseline$threshold, 1L, 0L)
cm_b <- table(Predicted = pred_class_baseline, Actual = actual_baseline)
print(cm_b)
cat("Sensitivity:", round(cm_b[2,2]/(cm_b[1,2]+cm_b[2,2]), 3), "\n")
cat("Specificity:", round(cm_b[1,1]/(cm_b[1,1]+cm_b[2,1]), 3), "\n")
cat("Precision:  ", round(cm_b[2,2]/(cm_b[2,1]+cm_b[2,2]), 3), "\n")
```

> **基准线：** Engineered model 必须在 AUC 上超越 `r round(auc(roc_baseline), 3)`。

---

## 6. Export

```{r export-analysis}
# Export tier mapping and vacancy keyword codes for use in 03_features.Rmd
building_tier_map %>%
  write_csv(paste0(out_path, "building_tier_mapping.csv"))

tibble(code = vacancy_keyword_codes) %>%
  write_csv(paste0(out_path, "vacancy_keyword_codes.csv"))

cat("已导出:\n")
cat("  building_tier_mapping.csv —", nrow(building_tier_map), "行\n")
cat("  vacancy_keyword_codes.csv —", length(vacancy_keyword_codes), "行\n")
```

---
